Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-24 numerical-ppm.tex
Edits to section 4.1 (PPM)

One (purely) hydrodynamic method used in \enzo\ is closely based on the
piecewise parabolic method (PPM) of
\citet{1984JCoPh..54..174C} -- henceforth referred to as CW84 -- which has been
modified for the study of cosmological fluid flows.  
CW84 describe two variants on this method:
Lagrangian Remap and Direct Eulerian; here we use the Direct Eulerian
version which is better suited for AMR simulations.  The Lagrangian
Remap version has previously been adapted for cosmological use
as described in \citet{1995CoPhC..89..149B}, and the implementation
we use here is closely based on that work.

The first term on the right hand side of equations~(\ref{eq:momentum}) and
(\ref{eq:total_energy}) comes from our choice of comoving coordinates
(a similar term also appears in equation~(\ref{eq:dm_velocity}), the velocity 
relation for the dark matter particles).  A similar term does not
appear in the mass conservation equation (\ref{eq:mass}) because
of the comoving density definition.   
We note that these expansion terms could be
eliminated entirely by the proper choice of variables (including time),
although we have not done so here, as they do not constitute a major
source of error.  We solve these terms
using the same technique -- they are split from the rest of the terms and
solved using an (implicit) time-centered method, which is straightforward
as there are no spatial gradients.  Note that we use this method for
all four of the hydro solvers.

@@

We now restrict ourselves to the solution of equations
(\ref{eq:mass1d})--(\ref{eq:total_energy1d}), in one
dimension.  In the Direct Eulerian version of PPM, this is
accomplished by a three-step process.  First, we compute effective
left and right states at each grid boundary by constructing a
piecewise parabolic description of the primative variables ($\rho$, $u$
and $E$) and then averaging over the regions corresponding to the
distance each characteristic wave can travel ($u$, $u-c_s$ and
$u+c_s$, where $c_s$ is the sound speed).
Second, a Riemann problem is solved using these effective left and
right states, and finally the fluxes are computed based on the
solution to this Riemann problem and the conserved quantities are
updated.  This is described in detail in
CW84, but we will briefly outline the procedure
here both for completeness and to put the changes we will make in
context.

@@

We first construct monotonic piecewise 
parabolic (third-order) interpolations in one dimension
for each of $p$, $\rho$, and $v$.  The pressure is determined from
equation~(\ref{eq:eq_of_state}), the equation of state.
The interpolation formula for some quantity $q$ is given by:
%
\begin{eqnarray}
q_j(x) & = &  q_{L,j} + \tilde{x}(\Delta q_j + q_{6,j}(1-\tilde{x})), \\
\tilde{x}      & \equiv & {x - x_{j-1/2} \over \Delta x_j}, \qquad
             x_{j-1/2} \leq x \leq x_{j+1/2}. \nonumber
\end{eqnarray}
%
This is equation~(1.4) of CW84 (in the spatial rather than mass coordinate as is appropriate for the direct Eulerian approach). 
The quantities $q_{L,j}$, $\Delta q_j$,
and $q_{6,j}$ can be viewed simply as interpolation constants; however,
they also have more intuitive meanings.  For example, $q_{L,j}$ is the
value of $q$ at the left edge of zone j, while $\Delta q_j$ and $q_{6,j}$ are analogous to the slope and first-order correction to the slope of $q$ (see CW84 for a complete discussion):
\begin{equation}
\Delta q_j \equiv q_{R,j} - q_{L,j} \qquad 
q_{6,j}    \equiv 6\left[q_j - 1/2\left(q_{L,j} + q_{R,j}\right)\right].
\end{equation}

@@

Once we have the reconstruction, the primary quantities ($p$, $\rho$, $v$ and $E$) are averaged over the domains corresponding to the three characteristics $C = u-c$, $u$ or $u+c$ (where $c$ is the sound speed in a cell):
\begin{equation}
      q^{C+}_{j+1/2} =
          {1 \over \Delta t C^n_j} \int^{x_{j+1/2}}_{x_{j+1/2}-\Delta t C^n_j} q_j(x) dx
\end{equation}
where $q^{C+}$ refers to the left characteristic $C$ (and a similar integral is done for the right characteristic $q^{C-}$).  The linearized gas dynamics equations are then used to compute second-order accurate left and right states that take into account the multiple wave families.  This process is described in CW84 and we use their equations (3.6) and (3.7).

@@

In addition, as discussed earlier, the three dimensional scheme is achieved by operator-zsplitting and repeating the above procedure in the other two orthogonal directions.  The transverse velocities and any additional passive quantities are naturally and easily added to this system (see Equation 3.6 of CW84).

@@

Large, hypersonic bulk flows appear to be very common in cosmological
simulations and they present a problem because of the high ratio of
kinetic energy $E_k$ to gas internal energy $e$, which can reach as
high as $10^8$.  Inverted, we see that the internal energy consists of
an extremely small portion of the total energy.  The pressure then,
proportional to $E - E_{k}$, is the small difference between two large
numbers: a disastrous numerical situation.  This is not as large a
problem as it may at first appear because it only occurs when the
pressure is negligibly small.  Therefore, even if we suffer large
errors in the pressure distribution in these regions, the dynamics and
total energy budget of the flow will remain unaffected.  Nevertheless,
if the temperature distribution is required for other reasons
(e.g., for calculating radiative processes), a remedy is required.

@@

It is necessary, however, to conserve the total energy so that the conversion
of kinetic to thermal energy is performed properly.   We must therefore,
combine the two formulations without allowing the separately advected
internal energy $e$ to play a role in the gas dynamics.  This is done by
carrying both terms through the simulation and using
the total energy $E$ for hydrodynamic routines and the internal energy
$e$ when the temperature profile is required.  One way to view this procedure
is to treat $e$ as enhanced precision (extra digits) for $E$ that automatically
`floats' to where it is needed.
We only require that they be kept synchronized when the two levels of precision
overlap.

@@

We note that others have independently developed a similar but
distinct scheme for dealing with this problem, which is endemic to
methods adopting the total energy equation.  In \citet{TVD93} the 
two variables adopted are total energy and
entropy (rather than total energy and thermal energy), with an
analogous scheme for choosing which variable to employ.

@@

While there exists an exact solution to this problem, finding it is expensive.  There are four
possible combination of left- and right- traveling shocks and rarefactions,
only one of which is fully consistent with the initial conditions.  Once the
correct physical state is determined, the pressure in the central region can
only be determined by finding the root to an algebraic equation, which is necessarily
an iterative process.  Thus a series of \emph{approximate} Riemann solvers are
typically used.  There are four approximate Riemann solvers in \enzo: two-shock
\citep{toro-1997},
Harten-Lax-van Leer \citep[HLL,][]{toro-1997}, HLL with a contact discontinuity
 \citep[HLLC,][]{toro-1997}, and HLL with
multiple discontinuities \citep[HLLD,][]{Miyoshi05}.  Two-shock is used only
with the PPM method.  HLL and HLLC are used with PPM, MUSCL (both with and
without MHD) and MHD-CT.  HLLD is exclusively an MHD solver, and works with both
MUSCL and MHD-CT methods.

The only approximation that two-shock makes is that both left- and right- moving
waves are shocks.  This solution still requires an iterative method for finding
the pressure in between the two waves.  The HLL method alleviates this iteration
by assuming that there is no central contact discontinuity, and the signal speed
in the central region is approximated by an average over the left- and right-
moving waves.  This method is significantly faster than the two-shock method,
but it is also quite a bit more dissipative.  The HLLC is a three-wave method that improves upon the HLL
method by also including the third wave, the contact discontinuity.

For the MHD equations, there are seven waves, instead of three.  This makes the
exact solution to the Riemann quite a bit more expensive.  Both the HLL and HLLC
approximations can be formulated for the MHD equations, and are employed in both
the Dedner and CT solvers in \enzo.  The HLLD
solver includes two of the additional waves, the rotational discontinuities,
making it a five-wave solver.

>>>>

One (purely) hydrodynamic method used in \enzo\ is closely based on the
piecewise parabolic method (PPM) of
\citet{1984JCoPh..54..174C} -- henceforth referred to as CW84 -- which has been
modified for the study of cosmological fluid flows.  
CW84 describe two variants on this method:
Lagrangian Remap and Direct Eulerian.  We use the Direct Eulerian
version in \enzo, which is better suited for AMR simulations.  The Lagrangian
Remap version has previously been adapted for cosmological use
as described in \citet{1995CoPhC..89..149B}, and the implementation
we use here is closely based on that work.

The first term on the right hand side of equations~(\ref{eq:momentum}) and
(\ref{eq:total_energy}) comes from our choice of comoving coordinates
(a similar term also appears in equation~(\ref{eq:dm_velocity}), the velocity 
relation for the dark matter particles).  A similar term does not
appear in the mass conservation equation (\ref{eq:mass}) because
of the comoving density definition.   
We note that these expansion terms could be
eliminated entirely by the proper choice of variables (including time),
although we have not done so here as they do not constitute a major
source of error.  We solve these terms
using the same technique -- they are split from the rest of the terms and
solved using an (implicit) time-centered method, which is straightforward
as there are no spatial gradients.  Note that we use this method for
all four of the hydro solvers.

@@

We now restrict ourselves to the solution of equations
(\ref{eq:mass1d})--(\ref{eq:total_energy1d}) in one
dimension.  In the Direct Eulerian version of PPM, this is
accomplished by a three-step process.  First, we compute effective
left and right states at each grid boundary by constructing a
piecewise parabolic description of the primative variables ($\rho$, $u$
and $E$) and then averaging over the regions corresponding to the
distance each characteristic wave can travel ($u$, $u-c_s$ and
$u+c_s$, where $c_s$ is the sound speed).
Second, a Riemann problem is solved using these effective left and
right states, and finally the fluxes are computed based on the
solution to this Riemann problem and the conserved quantities are
updated.  This is described in detail in
CW84, but we will briefly outline the procedure
here both for completeness and to put the changes we will make in
context.

@@

We first construct monotonic piecewise 
parabolic (third-order) interpolations in one dimension
for each of $p$, $\rho$, and $v$.  The pressure is determined from
equation~(\ref{eq:eq_of_state}), the equation of state.
The interpolation formula for some quantity $q$ is given by:
%
\begin{eqnarray}
q_j(x) & = &  q_{L,j} + \tilde{x}(\Delta q_j + q_{6,j}(1-\tilde{x})), \\
\tilde{x}      & \equiv & {x - x_{j-1/2} \over \Delta x_j}, \qquad
             x_{j-1/2} \leq x \leq x_{j+1/2}. \nonumber
\end{eqnarray}
%
This is equation~(1.4) of CW84 (in the spatial rather than mass coordinate, as is appropriate for the direct Eulerian approach). 
The quantities $q_{L,j}$, $\Delta q_j$,
and $q_{6,j}$ can be viewed simply as interpolation constants; however,
they also have more intuitive meanings.  For example, $q_{L,j}$ is the
value of $q$ at the left edge of zone j, while $\Delta q_j$ and $q_{6,j}$ are analogous to the slope and first-order correction to the slope of $q$ (see CW84 for a complete discussion):
\begin{equation}
\Delta q_j \equiv q_{R,j} - q_{L,j} \qquad 
q_{6,j}    \equiv 6\left[q_j - 1/2\left(q_{L,j} + q_{R,j}\right)\right].
\end{equation}

@@

Once we have the reconstruction, the primary quantities ($p$, $\rho$, $v$ and $E$) are averaged over the domains corresponding to the three characteristics $C = u-c$, $u$, or $u+c$ (where $c$ is the sound speed in a cell):
\begin{equation}
      q^{C+}_{j+1/2} =
          {1 \over \Delta t C^n_j} \int^{x_{j+1/2}}_{x_{j+1/2}-\Delta t C^n_j} q_j(x) dx
\end{equation}
where $q^{C+}$ refers to the left characteristic $C$ (and a similar integral is done for the right characteristic $q^{C-}$).  The linearized gas dynamics equations are then used to compute second-order accurate left and right states that take into account the multiple wave families.  This process is described in CW84 and we use their equations (3.6) and (3.7).

@@

In addition, as discussed earlier, the three-dimensional scheme is achieved by operator-splitting and repeating the above procedure in the other two orthogonal directions.  The transverse velocities and any additional passive quantities are naturally and easily added to this system (see Equation 3.6 of CW84).

@@

Large, hypersonic bulk flows appear to be very common in cosmological
simulations and they present a problem because of the high ratio of
kinetic energy $E_k$ to gas internal energy $e$, which can reach as
high as $10^8$.  Inverted, we see that the internal energy consists of
an extremely small portion of the total energy.  In such a situation, the pressure,
proportional to $E - E_{k}$, is the small difference between two large
numbers: a disastrous numerical situation.  This is not as large a
problem as it may at first appear because it only occurs when the
pressure is negligibly small.  Therefore, even if we suffer large
errors in the pressure distribution in these regions, the dynamics and
total energy budget of the flow will remain unaffected.  Nevertheless,
if the temperature distribution is required for other reasons
(e.g., for calculating radiative processes), a remedy is required.

@@

It is necessary, however, to conserve the total energy so that the conversion
of kinetic to thermal energy is performed properly.   We must therefore
combine the two formulations without allowing the separately-advected
internal energy $e$ to play a role in the gas dynamics.  This is done by
carrying both terms through the simulation and using
the total energy $E$ for hydrodynamic routines and the internal energy
$e$ when the temperature profile is required.  One way to view this procedure
is to treat $e$ as enhanced precision (extra digits) for $E$ that automatically
`floats' to where it is needed.
We only require that they be kept synchronized when the two levels of precision
overlap.

@@

We note that others have independently developed a similar but
distinct scheme for dealing with this problem, which is endemic to
methods adopting the total energy equation.  In \citet{TVD93}, the 
two variables adopted are total energy and
entropy (rather than total energy and thermal energy), with an
analogous scheme for choosing which variable to employ.

@@

While there exists an exact solution to this problem, finding it is expensive.  There are four
possible combination of left- and right- traveling shocks and rarefactions,
only one of which is fully consistent with the initial conditions.  Once the
correct physical state is determined, the pressure in the central region can
only be determined by finding the root to an algebraic equation, which is necessarily
an iterative process.  Thus a series of \emph{approximate} Riemann solvers are
typically used.  There are four approximate Riemann solvers in \enzo: two-shock
\citep{toro-1997},
Harten-Lax-van Leer \citep[HLL,][]{toro-1997}, HLL with a contact discontinuity
 \citep[HLLC,][]{toro-1997}, and HLL with
multiple discontinuities \citep[HLLD,][]{Miyoshi05}.  Two-shock is used only
with the PPM method.  HLL and HLLC are used with PPM, MUSCL (both with and
without MHD) and MHD-CT.  HLLD is exclusively an MHD solver, and works
with both the
MUSCL and MHD-CT methods.

The only approximation that two-shock makes is that both left- and right-moving
waves are shocks.  This solution still requires an iterative method for finding
the pressure in between the two waves.  The HLL method alleviates this iteration
by assuming that there is no central contact discontinuity, and the signal speed
in the central region is approximated by an average over the left- and right-
moving waves.  This method is significantly faster than the two-shock method,
but it is also quite a bit more dissipative.  The HLLC is a three-wave method that improves upon the HLL
method by also including the third wave, the contact discontinuity.

For the MHD equations, there are seven waves instead of three.  This makes the
exact solution to the Riemann problem quite a bit more expensive.  Both the HLL and HLLC
approximations can be formulated for the MHD equations, and are employed in both
the Dedner and CT solvers in \enzo.  The HLLD
solver includes two of the additional waves, the rotational discontinuities,
making it a five-wave solver.
